---
title: "Data Collection and Cleaning"
author: "Kyran Lakhia"
date: "`r Sys.Date()`"
output: pdf_document
---
#Initial Cleaning
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Packages
```{r}
library(readr)
library(dplyr)
library(tibble)
library(lubridate)
```

Read Raw Data
```{r}
games      <- read_csv("Games.csv", show_col_types = FALSE)
opp_pg     <- read_csv("Opponent Stats Per Game.csv", show_col_types = FALSE)
adv        <- read_csv("Advanced.csv", show_col_types = FALSE)
```

Inspect Column Names
```{r}
names(games)
names(opp_pg)
names(adv)
```

Historical Base Dataset
```{r}
games_master <- games %>%
  select(
    game_id   = gameId,
    date      = gameDateTimeEst,
    home_team = hometeamName,
    away_team = awayteamName,
    home_pts  = homeScore,
    away_pts  = awayScore
  ) %>%
  mutate(
    date   = as.Date(date),
    season = if_else(month(date) >= 10, year(date) + 1L, year(date)),
    spread = home_pts - away_pts,
    total  = home_pts + away_pts
  )
```

Team-Level Opponent Stats Dataset
```{r}
opp_team <- opp_pg %>%
  select(
    season,
    team,
    opp_fg_per_game,
    opp_fga_per_game,
    opp_fg_percent,
    opp_x3p_per_game,
    opp_x3pa_per_game,
    opp_x3p_percent,
    opp_x2p_per_game,
    opp_x2pa_per_game,
    opp_x2p_percent,
    opp_ft_per_game,
    opp_fta_per_game,
    opp_ft_percent,
    opp_orb_per_game,
    opp_drb_per_game,
    opp_trb_per_game,
    opp_ast_per_game,
    opp_stl_per_game,
    opp_blk_per_game,
    opp_tov_per_game,
    opp_pf_per_game,
    opp_pts_per_game
  ) %>%
  distinct()
```

Check Team Name Formats
```{r}
unique(games_master$home_team)[1:30]
unique(opp_team$team)[1:30]

sort(unique(games_master$season))[1:10]
sort(unique(opp_team$season))[1:10]
```
Team Name Table
```{r}
team_lookup <- tibble(
  short_team = c(
    "Hawks", "Celtics", "Nets", "Hornets", "Bulls", "Cavaliers",
    "Mavericks", "Nuggets", "Pistons", "Warriors", "Rockets", "Pacers",
    "Clippers", "Lakers", "Grizzlies", "Heat", "Bucks", "Timberwolves",
    "Pelicans", "Knicks", "Thunder", "Magic", "76ers", "Suns",
    "Trail Blazers", "Kings", "Spurs", "Raptors", "Jazz", "Wizards"
  ),
  full_team = c(
    "Atlanta Hawks", "Boston Celtics", "Brooklyn Nets", "Charlotte Hornets", "Chicago Bulls", "Cleveland Cavaliers",
    "Dallas Mavericks", "Denver Nuggets", "Detroit Pistons", "Golden State Warriors", "Houston Rockets", "Indiana Pacers",
    "Los Angeles Clippers", "Los Angeles Lakers", "Memphis Grizzlies", "Miami Heat", "Milwaukee Bucks", "Minnesota Timberwolves",
    "New Orleans Pelicans", "New York Knicks", "Oklahoma City Thunder", "Orlando Magic", "Philadelphia 76ers", "Phoenix Suns",
    "Portland Trail Blazers", "Sacramento Kings", "San Antonio Spurs", "Toronto Raptors", "Utah Jazz", "Washington Wizards"
  )
)
```

Map Names
```{r}
games_master <- games_master %>%
  left_join(team_lookup, by = c("home_team" = "short_team")) %>%
  rename(home_team_full = full_team) %>%
  left_join(team_lookup, by = c("away_team" = "short_team")) %>%
  rename(away_team_full = full_team)
```

Preparing Opponet Datasets
```{r}
home_opp <- opp_team %>%
  rename_with(~ paste0("home_", .x), -c(season, team))

away_opp <- opp_team %>%
  rename_with(~ paste0("away_", .x), -c(season, team))
```

Joining Opponet Stats onto Master Dataset
```{r}
games_master <- games_master %>%
  left_join(home_opp, by = c("season", "home_team_full" = "team")) %>%
  left_join(away_opp, by = c("season", "away_team_full" = "team"))
```

Creating Difference Variables
```{r}
games_master <- games_master %>%
  mutate(
    opp_pts_allowed_diff = home_opp_pts_per_game - away_opp_pts_per_game,
    opp_orb_allowed_diff = home_opp_orb_per_game - away_opp_orb_per_game,
    opp_trb_allowed_diff = home_opp_trb_per_game - away_opp_trb_per_game,
    opp_ast_allowed_diff = home_opp_ast_per_game - away_opp_ast_per_game,
    opp_tov_forced_diff  = home_opp_tov_per_game - away_opp_tov_per_game
  )
```

Filtering To Modern Era
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

Previewing Current Dataset
```{r}
games_master_modern %>%
  select(
    game_id, date, season, home_team, home_team_full, away_team, away_team_full,
    home_pts, away_pts, spread, total,
    home_opp_pts_per_game, away_opp_pts_per_game,
    opp_pts_allowed_diff, opp_orb_allowed_diff,
    opp_trb_allowed_diff, opp_ast_allowed_diff, opp_tov_forced_diff
  ) %>%
  head(20)
```

```{r}
write_csv(games_master, "master_dataset_v1.csv")
write_csv(games_master_modern, "master_dataset_modern_v1.csv")
```

#Creating OREB

```{r}
team_stats <- read_csv("TeamStatistics.csv", show_col_types = FALSE)
```

```{r}
names(team_stats)
```
Building Offensive Rebound Datasetr
```{r}
oreb_game <- team_stats %>%
  select(
    game_id = gameId,
    team = teamName,
    home = home,
    oreb = reboundsOffensive
  )
```

Splitting Into Home and Away
```{r}
# Home Offensive Rebounds

home_oreb <- oreb_game %>%
  filter(home == TRUE) %>%
  select(
    game_id,
    home_oreb = oreb
  )

# Away Offensive Rebounds

away_oreb <- oreb_game %>%
  filter(home == FALSE) %>%
  select(
    game_id,
    away_oreb = oreb
  )
```

Joining Onto Master
```{r}
## Join Offensive Rebounds onto Master Dataset

games_master <- games_master %>%
  left_join(home_oreb, by = "game_id") %>%
  left_join(away_oreb, by = "game_id") %>%
  mutate(
    oreb = home_oreb + away_oreb
  )
```

Verifying
```{r}
games_master %>%
  select(
    game_id,
    home_team,
    away_team,
    home_oreb,
    away_oreb,
    oreb
  ) %>%
  head(20)
```
Rebuilding Modern
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

Save
```{r}
write_csv(games_master, "master_dataset_v2.csv")
```
```{r}
write_csv(games_master_modern, "master_dataset_modern_v2.csv")
```

#Adding Advanced Team Stats

Advanced Team File
```{r}
team_adv <- read_csv("TeamStatisticsAdvanced.csv", show_col_types = FALSE)

names(team_adv)
head(team_adv)
```

Column-Matcher
```{r}
## Helper to find likely column names

pick_col <- function(df, candidates) {
  nms <- names(df)
  hit <- nms[tolower(nms) %in% tolower(candidates)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}
```

Identify Important Columns
```{r}
## Identify advanced-stat columns

game_id_col <- pick_col(team_adv, c("gameId"))
home_col    <- pick_col(team_adv, c("home"))

off_col  <- pick_col(team_adv, c("offensiveRating", "offRating", "off_rtg"))
def_col  <- pick_col(team_adv, c("defensiveRating", "defRating", "def_rtg"))
net_col  <- pick_col(team_adv, c("netRating", "net_rtg"))
pace_col <- pick_col(team_adv, c("pace"))
ts_col   <- pick_col(team_adv, c("trueShootingPercentage", "trueShootingPct", "tsPercent", "ts_percentage"))

game_id_col
home_col
off_col
def_col
net_col
pace_col
ts_col
```
Reduced
```{r}
team_adv_small <- team_adv %>%
  transmute(
    game_id = .data[[game_id_col]],
    home    = .data[[home_col]],
    off_rtg = .data[[off_col]],
    def_rtg = .data[[def_col]],
    net_rtg = .data[[net_col]],
    pace    = .data[[pace_col]]
  )
```

Game_ids cleaning
```{r}
games_master <- games_master %>%
  mutate(game_id = as.character(game_id))

team_adv_small <- team_adv %>%
  transmute(
    game_id = as.character(.data[[game_id_col]]),
    home    = .data[[home_col]],
    off_rtg = .data[[off_col]],
    def_rtg = .data[[def_col]],
    net_rtg = .data[[net_col]],
    pace    = .data[[pace_col]]
  )
```

Splitting Home vs. Away
```{r}
home_adv <- team_adv_small %>%
  filter(home == 1) %>%
  select(
    game_id,
    home_off_rtg = off_rtg,
    home_def_rtg = def_rtg,
    home_net_rtg = net_rtg,
    home_pace    = pace
  ) %>%
  distinct(game_id, .keep_all = TRUE)

away_adv <- team_adv_small %>%
  filter(home == 0) %>%
  select(
    game_id,
    away_off_rtg = off_rtg,
    away_def_rtg = def_rtg,
    away_net_rtg = net_rtg,
    away_pace    = pace
  ) %>%
  distinct(game_id, .keep_all = TRUE)
```

Rejoining
```{r}
games_master <- games_master %>%
  left_join(home_adv, by = "game_id") %>%
  left_join(away_adv, by = "game_id") %>%
  mutate(
    off_rtg_diff = home_off_rtg - away_off_rtg,
    def_rtg_diff = home_def_rtg - away_def_rtg,
    net_rtg_diff = home_net_rtg - away_net_rtg,
    pace_diff    = home_pace - away_pace
  )
```

IDs Check
```{r}
class(games_master$game_id)
class(team_adv_small$game_id)

length(intersect(games_master$game_id, team_adv_small$game_id))
sum(games_master$game_id %in% team_adv_small$game_id)
```

viewing master
```{r}
games_master %>%
  filter(!is.na(home_off_rtg)) %>%
  select(
    game_id, home_team, away_team,
    home_off_rtg, away_off_rtg, off_rtg_diff,
    home_def_rtg, away_def_rtg, def_rtg_diff,
    home_net_rtg, away_net_rtg, net_rtg_diff,
    home_pace, away_pace, pace_diff
  ) %>%
  head(20)
```
#Adding Four Factors

Loading Dataset
```{r}
team_ff <- read_csv("TeamStatisticsFourFactors.csv", show_col_types = FALSE)
```

Inspecting Columns
```{r}
names(team_ff)
head(team_ff)
```

Identifying Columns
```{r}
ff_game_id_col <- pick_col(team_ff, c("gameId"))
ff_home_col    <- pick_col(team_ff, c("home"))

efg_col <- pick_col(team_ff, c("efgPct"))
tov_col <- pick_col(team_ff, c("tmTOVPct"))
orb_col <- pick_col(team_ff, c("orebPct"))
ftr_col <- pick_col(team_ff, c("ftaRate"))

ff_game_id_col
ff_home_col
efg_col
tov_col
orb_col
ftr_col
```
Building Reduced Table
```{r}
team_ff_small <- team_ff %>%
  transmute(
    game_id = as.character(.data[[ff_game_id_col]]),
    home    = .data[[ff_home_col]],
    efg_pct = .data[[efg_col]],
    tov_pct = .data[[tov_col]],
    orb_pct = .data[[orb_col]],
    ft_rate = .data[[ftr_col]]
  )
```

Splitting Into Home and Away
```{r}
home_ff <- team_ff_small %>%
  filter(home == 1) %>%
  select(
    game_id,
    home_efg_pct = efg_pct,
    home_tov_pct = tov_pct,
    home_orb_pct = orb_pct,
    home_ft_rate = ft_rate
  ) %>%
  distinct(game_id, .keep_all = TRUE)

away_ff <- team_ff_small %>%
  filter(home == 0) %>%
  select(
    game_id,
    away_efg_pct = efg_pct,
    away_tov_pct = tov_pct,
    away_orb_pct = orb_pct,
    away_ft_rate = ft_rate
  ) %>%
  distinct(game_id, .keep_all = TRUE)
```

Join onto master
```{r}
games_master <- games_master %>%
  mutate(game_id = as.character(game_id)) %>%
  left_join(home_ff, by = "game_id") %>%
  left_join(away_ff, by = "game_id") %>%
  mutate(
    efg_pct_diff = home_efg_pct - away_efg_pct,
    tov_pct_diff = home_tov_pct - away_tov_pct,
    orb_pct_diff = home_orb_pct - away_orb_pct,
    ft_rate_diff = home_ft_rate - away_ft_rate
  )
```

Verifying
```{r}
games_master %>%
  filter(!is.na(home_efg_pct)) %>%
  select(
    game_id, home_team, away_team,
    home_efg_pct, away_efg_pct, efg_pct_diff,
    home_tov_pct, away_tov_pct, tov_pct_diff,
    home_orb_pct, away_orb_pct, orb_pct_diff,
    home_ft_rate, away_ft_rate, ft_rate_diff
  ) %>%
  head(20)
```
Checking For Overlap
```{r}
length(intersect(games_master$game_id, team_ff_small$game_id))

sum(is.na(games_master$home_efg_pct))
sum(is.na(games_master$away_efg_pct))
sum(is.na(games_master$efg_pct_diff))
```
Filtering
```{r}
games_master_modern <- games_master %>%
  filter(!is.na(home_off_rtg)) %>%
  filter(!is.na(home_efg_pct))
```

Saving
```{r}
write_csv(games_master, "master_dataset_v4.csv")
write_csv(games_master_modern, "master_dataset_modern_v4.csv")
```

#Rolling Averages

Packages
```{r}
library(zoo)
```

Team-Level Game History
```{r}
team_games_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team,
    pts = home_pts,
    pts_allowed = away_pts,
    oreb = home_oreb
  )

team_games_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team,
    pts = away_pts,
    pts_allowed = home_pts,
    oreb = away_oreb
  )

team_games <- bind_rows(team_games_home, team_games_away) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date, game_id)
```

Computing Rolling Features
```{r}
team_games <- team_games %>%
  group_by(team) %>%
  arrange(date, game_id, .by_group = TRUE) %>%
  mutate(
    last5_pts = rollmean(lag(pts), 5, fill = NA, align = "right"),
    last5_pts_allowed = rollmean(lag(pts_allowed), 5, fill = NA, align = "right"),
    last5_oreb = rollmean(lag(oreb), 5, fill = NA, align = "right")
  ) %>%
  ungroup()
```

Home Rolling Stats
```{r}
home_recent <- team_games %>%
  select(
    game_id,
    team,
    last5_pts,
    last5_pts_allowed,
    last5_oreb
  ) %>%
  rename(
    home_team = team,
    home_last5_pts = last5_pts,
    home_last5_pts_allowed = last5_pts_allowed,
    home_last5_oreb = last5_oreb
  )
```

Away Rolling Stats
```{r}
away_recent <- team_games %>%
  select(
    game_id,
    team,
    last5_pts,
    last5_pts_allowed,
    last5_oreb
  ) %>%
  rename(
    away_team = team,
    away_last5_pts = last5_pts,
    away_last5_pts_allowed = last5_pts_allowed,
    away_last5_oreb = last5_oreb
  )
```

Joining with Master
```{r}
games_master <- games_master %>%
  select(-any_of(c(
    "home_last5_pts", "away_last5_pts", "last5_pts_diff",
    "home_last5_pts_allowed", "away_last5_pts_allowed", "last5_pts_allowed_diff",
    "home_last5_oreb", "away_last5_oreb", "last5_oreb_diff"
  ))) %>%
  left_join(home_recent, by = c("game_id", "home_team")) %>%
  left_join(away_recent, by = c("game_id", "away_team")) %>%
  mutate(
    last5_pts_diff = home_last5_pts - away_last5_pts,
    last5_pts_allowed_diff = home_last5_pts_allowed - away_last5_pts_allowed,
    last5_oreb_diff = home_last5_oreb - away_last5_oreb
  )
```

Verifying
```{r}
games_master %>%
  select(
    game_id, date, home_team, away_team,
    home_last5_pts, away_last5_pts, last5_pts_diff,
    home_last5_pts_allowed, away_last5_pts_allowed, last5_pts_allowed_diff,
    home_last5_oreb, away_last5_oreb, last5_oreb_diff
  ) %>%
  filter(!is.na(home_last5_pts) | !is.na(away_last5_pts)) %>%
  head(20)
```

Rebuild Modern
```{r}
games_master_modern <- games_master %>%
  filter(!is.na(home_off_rtg)) %>%
  filter(!is.na(home_efg_pct))
```

Save
```{r}
write_csv(games_master, "master_dataset_v5.csv")
write_csv(games_master_modern, "master_dataset_modern_v5.csv")
```

#Schedule Factors
Building Team Level Schedule
```{r}
team_games_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team
  )

team_games_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team
  )

team_games <- bind_rows(team_games_home, team_games_away) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date, game_id)
```

Days Since Last Game
```{r}
team_games <- team_games %>%
  group_by(team) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    rest_days = as.numeric(date - lag(date))
  ) %>%
  ungroup()
```

Back-to-Back Indictator
```{r}
team_games <- team_games %>%
  mutate(
    back_to_back = if_else(rest_days == 1, 1, 0)
  )
```

Split Home vs Away Rest
```{r}
home_rest <- team_games %>%
  rename(home_team = team) %>%
  select(
    game_id,
    home_team,
    home_rest_days = rest_days,
    home_back_to_back = back_to_back
  )

away_rest <- team_games %>%
  rename(away_team = team) %>%
  select(
    game_id,
    away_team,
    away_rest_days = rest_days,
    away_back_to_back = back_to_back
  )
```

Join Onto Master
```{r}
games_master <- games_master %>%
  left_join(home_rest, by = c("game_id", "home_team")) %>%
  left_join(away_rest, by = c("game_id", "away_team")) %>%
  mutate(
    rest_diff = home_rest_days - away_rest_days
  )
```

Rebuild Modern
```{r}
games_master_modern <- games_master %>%
  filter(!is.na(home_off_rtg)) %>%
  filter(!is.na(home_efg_pct))
```

Save
```{r}
write_csv(games_master, "master_dataset_v6.csv")
write_csv(games_master_modern, "master_dataset_modern_v6.csv")
```

Verify
```{r}
games_master %>%
  select(
    game_id,
    home_team,
    away_team,
    home_rest_days,
    away_rest_days,
    rest_diff,
    home_back_to_back,
    away_back_to_back
  ) %>%
  head(20)
```

#Rolling Team Strength

Team-Level Advanced History
```{r}
team_adv_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team,
    off_rtg = home_off_rtg,
    def_rtg = home_def_rtg,
    net_rtg = home_net_rtg,
    pace = home_pace
  )

team_adv_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team,
    off_rtg = away_off_rtg,
    def_rtg = away_def_rtg,
    net_rtg = away_net_rtg,
    pace = away_pace
  )

team_adv_games <- bind_rows(team_adv_home, team_adv_away) %>%
  arrange(team, date)
```

Computing Rolling Averages
```{r}
team_adv_games <- team_adv_games %>%
  group_by(team) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    last5_off_rtg = zoo::rollapply(lag(off_rtg), 5, mean, fill = NA, align = "right"),
    last5_def_rtg = zoo::rollapply(lag(def_rtg), 5, mean, fill = NA, align = "right"),
    last5_net_rtg = zoo::rollapply(lag(net_rtg), 5, mean, fill = NA, align = "right"),
    last5_pace    = zoo::rollapply(lag(pace), 5, mean, fill = NA, align = "right")
  ) %>%
  ungroup()
```

Split Home vs. Away
```{r}
home_recent_adv <- team_adv_games %>%
  rename(home_team = team) %>%
  select(
    game_id,
    home_team,
    home_last5_off_rtg = last5_off_rtg,
    home_last5_def_rtg = last5_def_rtg,
    home_last5_net_rtg = last5_net_rtg,
    home_last5_pace = last5_pace
  )

away_recent_adv <- team_adv_games %>%
  rename(away_team = team) %>%
  select(
    game_id,
    away_team,
    away_last5_off_rtg = last5_off_rtg,
    away_last5_def_rtg = last5_def_rtg,
    away_last5_net_rtg = last5_net_rtg,
    away_last5_pace = last5_pace
  )
```

Join Onto Master
```{r}
games_master <- games_master %>%
  left_join(home_recent_adv, by = c("game_id", "home_team")) %>%
  left_join(away_recent_adv, by = c("game_id", "away_team")) %>%
  mutate(
    last5_off_rtg_diff = home_last5_off_rtg - away_last5_off_rtg,
    last5_def_rtg_diff = home_last5_def_rtg - away_last5_def_rtg,
    last5_net_rtg_diff = home_last5_net_rtg - away_last5_net_rtg,
    last5_pace_diff = home_last5_pace - away_last5_pace
  )
```

Rebuild Modern
```{r}
games_master_modern <- games_master %>%
  filter(!is.na(home_off_rtg)) %>%
  filter(!is.na(home_efg_pct))
```

Save
```{r}
write_csv(games_master, "master_dataset_v7.csv")
write_csv(games_master_modern, "master_dataset_modern_v7.csv")
```

#Final Edits
```{r}
names(games_master_modern)
```
Selecting Variabels
```{r}
model_data <- games_master_modern %>%
  select(
    game_id,
    date,
    home_team,
    away_team,

    spread,
    total,
    oreb,

    off_rtg_diff,
    def_rtg_diff,
    net_rtg_diff,
    pace_diff,

    efg_pct_diff,
    tov_pct_diff,
    orb_pct_diff,
    ft_rate_diff,

    opp_pts_allowed_diff,
    opp_orb_allowed_diff,
    opp_trb_allowed_diff,
    opp_ast_allowed_diff,
    opp_tov_forced_diff,

    last5_pts_diff,
    last5_pts_allowed_diff,
    last5_oreb_diff,

    last5_off_rtg_diff,
    last5_def_rtg_diff,
    last5_net_rtg_diff,
    last5_pace_diff,

    rest_diff,
    home_back_to_back,
    away_back_to_back
  )
```

Save
```{r}
write_csv(model_data, "FINAL_CLEANED_SET.csv")
```
