---
title: "Data Collection and Cleaning"
author: "Kyran Lakhia"
date: "`r Sys.Date()`"
output: pdf_document
---
##Initial Cleaning
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Load Packages
```{r}
library(readr)
library(dplyr)
library(tibble)
library(lubridate)
library(zoo)
library(tidyr)
library(slider)
```

```{r}
pick_col <- function(df, candidates) {
  nms <- names(df)
  hit <- nms[tolower(nms) %in% tolower(candidates)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}
```
Read Raw Data
```{r}
games <- read_csv("Data/Games.csv", show_col_types = FALSE)
opp_pg <- read_csv("Data/Opponent Stats Per Game.csv", show_col_types = FALSE)
```

Inspect Column Names
```{r}
names(games)
names(opp_pg)
```

Historical Base Dataset
```{r}
games_master <- games %>%
  select(
    game_id   = gameId,
    date      = gameDateTimeEst,
    home_team = hometeamName,
    away_team = awayteamName,
    home_pts  = homeScore,
    away_pts  = awayScore
  ) %>%
  mutate(
    date   = as.Date(date),
    season = if_else(month(date) >= 10, year(date) + 1L, year(date)),
    spread = home_pts - away_pts,
    total  = home_pts + away_pts
  )
```

Team-Level Opponent Stats Dataset
```{r}
opp_team <- opp_pg %>%
  select(
    season,
    team,
    opp_fg_per_game,
    opp_fga_per_game,
    opp_fg_percent,
    opp_x3p_per_game,
    opp_x3pa_per_game,
    opp_x3p_percent,
    opp_x2p_per_game,
    opp_x2pa_per_game,
    opp_x2p_percent,
    opp_ft_per_game,
    opp_fta_per_game,
    opp_ft_percent,
    opp_orb_per_game,
    opp_drb_per_game,
    opp_trb_per_game,
    opp_ast_per_game,
    opp_stl_per_game,
    opp_blk_per_game,
    opp_tov_per_game,
    opp_pf_per_game,
    opp_pts_per_game
  ) %>%
  distinct()
```

Check Team Name Formats
```{r}
unique(games_master$home_team)[1:30]
unique(opp_team$team)[1:30]

sort(unique(games_master$season))[1:10]
sort(unique(opp_team$season))[1:10]
```
Team Name Table
```{r}
team_lookup <- tibble(
  short_team = c(
    "Hawks", "Celtics", "Nets", "Hornets", "Bulls", "Cavaliers",
    "Mavericks", "Nuggets", "Pistons", "Warriors", "Rockets", "Pacers",
    "Clippers", "Lakers", "Grizzlies", "Heat", "Bucks", "Timberwolves",
    "Pelicans", "Knicks", "Thunder", "Magic", "76ers", "Suns",
    "Trail Blazers", "Kings", "Spurs", "Raptors", "Jazz", "Wizards"
  ),
  full_team = c(
    "Atlanta Hawks", "Boston Celtics", "Brooklyn Nets", "Charlotte Hornets", "Chicago Bulls", "Cleveland Cavaliers",
    "Dallas Mavericks", "Denver Nuggets", "Detroit Pistons", "Golden State Warriors", "Houston Rockets", "Indiana Pacers",
    "Los Angeles Clippers", "Los Angeles Lakers", "Memphis Grizzlies", "Miami Heat", "Milwaukee Bucks", "Minnesota Timberwolves",
    "New Orleans Pelicans", "New York Knicks", "Oklahoma City Thunder", "Orlando Magic", "Philadelphia 76ers", "Phoenix Suns",
    "Portland Trail Blazers", "Sacramento Kings", "San Antonio Spurs", "Toronto Raptors", "Utah Jazz", "Washington Wizards"
  )
)
```

Map Names
```{r}
games_master <- games_master %>%
  left_join(team_lookup, by = c("home_team" = "short_team")) %>%
  rename(home_team_full = full_team) %>%
  left_join(team_lookup, by = c("away_team" = "short_team")) %>%
  rename(away_team_full = full_team)
```

Preparing Opponet Datasets
```{r}
home_opp <- opp_team %>%
  rename_with(~ paste0("home_", .x), -c(season, team))

away_opp <- opp_team %>%
  rename_with(~ paste0("away_", .x), -c(season, team))
```

Joining Opponet Stats onto Master Dataset
```{r}
games_master <- games_master %>%
  left_join(home_opp, by = c("season", "home_team_full" = "team")) %>%
  left_join(away_opp, by = c("season", "away_team_full" = "team"))
```

Creating Difference Variables
```{r}
games_master <- games_master %>%
  mutate(
    opp_pts_allowed_diff = home_opp_pts_per_game - away_opp_pts_per_game,
    opp_orb_allowed_diff = home_opp_orb_per_game - away_opp_orb_per_game,
    opp_trb_allowed_diff = home_opp_trb_per_game - away_opp_trb_per_game,
    opp_ast_allowed_diff = home_opp_ast_per_game - away_opp_ast_per_game,
    opp_tov_forced_diff  = home_opp_tov_per_game - away_opp_tov_per_game
  )
```

Filtering To Modern Era
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

Previewing Current Dataset
```{r}
games_master_modern %>%
  select(
    game_id, date, season, home_team, home_team_full, away_team, away_team_full,
    home_pts, away_pts, spread, total,
    home_opp_pts_per_game, away_opp_pts_per_game,
    opp_pts_allowed_diff, opp_orb_allowed_diff,
    opp_trb_allowed_diff, opp_ast_allowed_diff, opp_tov_forced_diff
  ) %>%
  head(20)
```

##Creating OREB

```{r}
team_stats <- read_csv("Data/TeamStatistics.csv", show_col_types = FALSE)
```

```{r}
names(team_stats)
```
```{r}
## ------------------------------------------------------------
## ENGINEER LEVIN VARIABLES FROM TEAM BOX SCORE DATA
## ------------------------------------------------------------

team_box <- team_stats %>%
  transmute(
    game_id = as.character(gameId),
    team = teamName,
    home = home,

    pts = teamScore,
    fgm = fieldGoalsMade,
    fga = fieldGoalsAttempted,
    x3pm = threePointersMade,
    x3pa = threePointersAttempted,
    ftm = freeThrowsMade,
    fta = freeThrowsAttempted,
    oreb = reboundsOffensive,
    dreb = reboundsDefensive,
    tov = turnovers
  )

## --------------------------
## Compute base efficiencies
## --------------------------

team_box <- team_box %>%
  mutate(

    # Effective FG%
    efg_pct = (fgm + 0.5 * x3pm) / fga,

    # FT rate
    ft_rate = fta / fga,

    # Turnover %
    tov_pct = tov / (fga + 0.44 * fta + tov),

    # Possessions estimate
    possessions =
      fga -
      oreb +
      tov +
      0.44 * fta
  )

## --------------------------
## Join opponent stats
## --------------------------

opp_lookup <- team_box %>%
  select(game_id, team, pts, possessions, dreb)

team_box <- team_box %>%
  inner_join(
    opp_lookup,
    by = "game_id",
    suffix = c("", "_opp")
  ) %>%
  filter(team != team_opp)

## --------------------------
## Defensive stats
## --------------------------

team_box <- team_box %>%
  mutate(

    # Offensive rating
    off_rtg = 100 * pts / possessions,

    # Defensive rating
    def_rtg = 100 * pts_opp / possessions_opp,

    # Net rating
    net_rtg = off_rtg - def_rtg,

    # Pace estimate
    pace = (possessions + possessions_opp) / 2,

    # Offensive rebound %
    orb_pct = oreb / (oreb + dreb_opp)
  )

## --------------------------
## Rolling Levin metrics
## --------------------------

games_master <- games_master %>%
  mutate(game_id = as.character(game_id))

team_box <- team_box %>%
  left_join(
    games_master %>%
      select(game_id, date),
    by = "game_id"
  ) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date)

team_box <- team_box %>%
  group_by(team) %>%
  arrange(date, game_id, .by_group = TRUE) %>%
  mutate(

    last5_off_rtg =
      zoo::rollapply(lag(off_rtg), 5, mean, fill = NA, align = "right"),

    last5_def_rtg =
      zoo::rollapply(lag(def_rtg), 5, mean, fill = NA, align = "right"),

    last5_net_rtg =
      zoo::rollapply(lag(net_rtg), 5, mean, fill = NA, align = "right"),

    last5_pace =
      zoo::rollapply(lag(pace), 5, mean, fill = NA, align = "right")

  ) %>%
  ungroup()

## --------------------------
## Split home / away
## --------------------------

home_levin <- team_box %>%
  filter(home == 1) %>%
  select(
    game_id,
    home_team = team,
    home_off_rtg = off_rtg,
    home_def_rtg = def_rtg,
    home_net_rtg = net_rtg,
    home_pace = pace,
    home_efg_pct = efg_pct,
    home_tov_pct = tov_pct,
    home_orb_pct = orb_pct,
    home_ft_rate = ft_rate,
    home_last5_off_rtg = last5_off_rtg,
    home_last5_def_rtg = last5_def_rtg,
    home_last5_net_rtg = last5_net_rtg,
    home_last5_pace = last5_pace
  ) %>%
  distinct(game_id, home_team, .keep_all = TRUE)

away_levin <- team_box %>%
  filter(home == 0) %>%
  select(
    game_id,
    away_team = team,
    away_off_rtg = off_rtg,
    away_def_rtg = def_rtg,
    away_net_rtg = net_rtg,
    away_pace = pace,
    away_efg_pct = efg_pct,
    away_tov_pct = tov_pct,
    away_orb_pct = orb_pct,
    away_ft_rate = ft_rate,
    away_last5_off_rtg = last5_off_rtg,
    away_last5_def_rtg = last5_def_rtg,
    away_last5_net_rtg = last5_net_rtg,
    away_last5_pace = last5_pace
  ) %>%
  distinct(game_id, away_team, .keep_all = TRUE)

## --------------------------
## Join into games_master
## --------------------------

games_master <- games_master %>%
  left_join(home_levin, by = c("game_id", "home_team")) %>%
  left_join(away_levin, by = c("game_id", "away_team")) %>%
  mutate(

    off_rtg_diff = home_off_rtg - away_off_rtg,
    def_rtg_diff = home_def_rtg - away_def_rtg,
    net_rtg_diff = home_net_rtg - away_net_rtg,
    pace_diff = home_pace - away_pace,

    efg_pct_diff = home_efg_pct - away_efg_pct,
    tov_pct_diff = home_tov_pct - away_tov_pct,
    orb_pct_diff = home_orb_pct - away_orb_pct,
    ft_rate_diff = home_ft_rate - away_ft_rate,

    last5_off_rtg_diff =
      home_last5_off_rtg - away_last5_off_rtg,

    last5_def_rtg_diff =
      home_last5_def_rtg - away_last5_def_rtg,

    last5_net_rtg_diff =
      home_last5_net_rtg - away_last5_net_rtg,

    last5_pace_diff =
      home_last5_pace - away_last5_pace
  )
```

Building Offensive Rebound Dataset
```{r}
oreb_game <- team_stats %>%
  select(
    game_id = gameId,
    team = teamName,
    home = home,
    oreb = reboundsOffensive
  )
```

Splitting Into Home and Away
```{r}
# Home Offensive Rebounds

home_oreb <- oreb_game %>%
  filter(home == 1) %>%
  select(
    game_id,
    home_team = team,
    home_oreb = oreb
  ) %>%
  mutate(game_id = as.character(game_id))

away_oreb <- oreb_game %>%
  filter(home == 0) %>%
  select(
    game_id,
    away_team = team,
    away_oreb = oreb
  ) %>%
  mutate(game_id = as.character(game_id))
```

Joining Onto Master
```{r}
## Join Offensive Rebounds onto Master Dataset

games_master <- games_master %>%
  select(-any_of(c("home_oreb", "away_oreb", "oreb"))) %>%
  mutate(game_id = as.character(game_id)) %>%
  left_join(home_oreb, by = c("game_id", "home_team")) %>%
  left_join(away_oreb, by = c("game_id", "away_team")) %>%
  mutate(
    oreb = home_oreb + away_oreb
  )
```

Verifying
```{r}
games_master %>%
  select(
    game_id,
    home_team,
    away_team,
    home_oreb,
    away_oreb,
    oreb
  ) %>%
  head(20)
```
Rebuilding Modern
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```


##Rolling Averages

Packages
```{r}
library(zoo)
```

Team-Level Game History
```{r}
team_games_roll_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team,
    pts = home_pts,
    pts_allowed = away_pts,
    oreb = home_oreb
  )

team_games_roll_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team,
    pts = away_pts,
    pts_allowed = home_pts,
    oreb = away_oreb
  )

team_games_roll <- bind_rows(team_games_roll_home, team_games_roll_away) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date, game_id)
```

Computing Rolling Features
```{r}
team_games_roll <- team_games_roll %>%
  group_by(team) %>%
  arrange(date, game_id, .by_group = TRUE) %>%
  mutate(
    last5_pts = rollmean(lag(pts), 5, fill = NA, align = "right"),
    last5_pts_allowed = rollmean(lag(pts_allowed), 5, fill = NA, align = "right"),
    last5_oreb = rollmean(lag(oreb), 5, fill = NA, align = "right")
  ) %>%
  ungroup()
```

Home Rolling Stats
```{r}
home_recent <- team_games_roll %>%
  select(
    game_id,
    team,
    last5_pts,
    last5_pts_allowed,
    last5_oreb
  ) %>%
  rename(
    home_team = team,
    home_last5_pts = last5_pts,
    home_last5_pts_allowed = last5_pts_allowed,
    home_last5_oreb = last5_oreb
  )
```

Away Rolling Stats
```{r}
away_recent_roll <- team_games_roll %>%
  select(
    game_id,
    team,
    last5_pts,
    last5_pts_allowed,
    last5_oreb
  ) %>%
  rename(
    away_team = team,
    away_last5_pts = last5_pts,
    away_last5_pts_allowed = last5_pts_allowed,
    away_last5_oreb = last5_oreb
  )
```

Joining with Master
```{r}
games_master <- games_master %>%
  select(-any_of(c(
    "home_last5_pts", "away_last5_pts", "last5_pts_diff",
    "home_last5_pts_allowed", "away_last5_pts_allowed", "last5_pts_allowed_diff",
    "home_last5_oreb", "away_last5_oreb", "last5_oreb_diff"
  ))) %>%
  left_join(home_recent, by = c("game_id", "home_team")) %>%
  left_join(away_recent_roll, by = c("game_id", "away_team")) %>%
  mutate(
    last5_pts_diff = home_last5_pts - away_last5_pts,
    last5_pts_allowed_diff = home_last5_pts_allowed - away_last5_pts_allowed,
    last5_oreb_diff = home_last5_oreb - away_last5_oreb
  )
```

Verifying
```{r}
games_master %>%
  select(
    game_id, date, home_team, away_team,
    home_last5_pts, away_last5_pts, last5_pts_diff,
    home_last5_pts_allowed, away_last5_pts_allowed, last5_pts_allowed_diff,
    home_last5_oreb, away_last5_oreb, last5_oreb_diff
  ) %>%
  filter(!is.na(home_last5_pts) | !is.na(away_last5_pts)) %>%
  head(20)
```

Rebuild Modern
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

##Schedule Factors
Building Team Level Schedule
```{r}
team_games_rest_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team
  )

team_games_rest_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team
  )

team_games_rest <- bind_rows(team_games_rest_home, team_games_rest_away) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date, game_id)
```

Days Since Last Game
```{r}
team_games_rest <- team_games_rest %>%
  group_by(team) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    rest_days = as.numeric(date - lag(date))
  ) %>%
  ungroup()
```

Back-to-Back Indictator
```{r}
team_games_rest <- team_games_rest %>%
  mutate(
    back_to_back = if_else(rest_days == 1, 1, 0)
  )
```

Split Home vs Away Rest
```{r}
home_rest <- team_games_rest %>%
  rename(home_team = team) %>%
  select(
    game_id,
    home_team,
    home_rest_days = rest_days,
    home_back_to_back = back_to_back
  )

away_rest <- team_games_rest %>%
  rename(away_team = team) %>%
  select(
    game_id,
    away_team,
    away_rest_days = rest_days,
    away_back_to_back = back_to_back
  )
```

Join Onto Master
```{r}
games_master <- games_master %>%
  left_join(home_rest, by = c("game_id", "home_team")) %>%
  left_join(away_rest, by = c("game_id", "away_team")) %>%
  mutate(
    rest_diff = home_rest_days - away_rest_days
  )
```

Rebuild Modern
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

Verify
```{r}
games_master %>%
  select(
    game_id,
    home_team,
    away_team,
    home_rest_days,
    away_rest_days,
    rest_diff,
    home_back_to_back,
    away_back_to_back
  ) %>%
  head(20)
```

##Fixes

Removing Preseason
```{r}
games_master <- games_master %>%
  filter(!(month(date) == 10 & day(date) < 18))
```

Only NBA
```{r}
nba_teams <- team_lookup$short_team

games_master <- games_master %>%
  filter(home_team %in% nba_teams,
         away_team %in% nba_teams)
```

Removing games w/o full history
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010) %>%
  filter(!is.na(home_last5_pts), !is.na(away_last5_pts))
```

##OREB Work



##OREB Improvement Features (future-safe)


Read additional team-level season files
```{r}
team_pg <- read_csv("Data/Team Stats Per Game.csv", show_col_types = FALSE)
team_100 <- read_csv("Data/Team Stats Per 100 Poss.csv", show_col_types = FALSE)
```


#Team Stats Per Game: season-level opportunity/rebound features
```{r}
pg_team_col   <- pick_col(team_pg, c("team", "team_name", "teamName"))
pg_season_col <- pick_col(team_pg, c("season"))
pg_fg_col     <- pick_col(team_pg, c("fg_per_game", "fg"))
pg_fga_col    <- pick_col(team_pg, c("fga_per_game", "fga"))
pg_x3pa_col   <- pick_col(team_pg, c("x3pa_per_game", "x3pa", "3pa_per_game"))
pg_orb_col    <- pick_col(team_pg, c("orb_per_game", "oreb_per_game", "orb"))
pg_drb_col    <- pick_col(team_pg, c("drb_per_game", "dreb_per_game", "drb"))
pg_trb_col    <- pick_col(team_pg, c("trb_per_game", "reb_per_game", "trb"))

pg_team_col
pg_season_col
pg_fg_col
pg_fga_col
pg_x3pa_col
pg_orb_col
pg_drb_col
pg_trb_col

team_pg_small <- team_pg %>%
  transmute(
    season = .data[[pg_season_col]],
    team   = .data[[pg_team_col]],
    fg_pg  = .data[[pg_fg_col]],
    fga_pg = .data[[pg_fga_col]],
    x3pa_pg = .data[[pg_x3pa_col]],
    orb_pg = .data[[pg_orb_col]],
    drb_pg = .data[[pg_drb_col]],
    trb_pg = .data[[pg_trb_col]]
  ) %>%
  mutate(
    missed_fg_pg = fga_pg - fg_pg,
    x3pa_rate_pg = if_else(fga_pg > 0, x3pa_pg / fga_pg, NA_real_)
  ) %>%
  distinct()

home_pg <- team_pg_small %>%
  rename_with(~ paste0("home_", .x), -c(season, team))

away_pg <- team_pg_small %>%
  rename_with(~ paste0("away_", .x), -c(season, team))

games_master <- games_master %>%
  left_join(home_pg, by = c("season", "home_team_full" = "team")) %>%
  left_join(away_pg, by = c("season", "away_team_full" = "team")) %>%
  mutate(
    orb_pg_diff        = home_orb_pg - away_orb_pg,
    drb_pg_diff        = home_drb_pg - away_drb_pg,
    trb_pg_diff        = home_trb_pg - away_trb_pg,
    missed_fg_pg_diff  = home_missed_fg_pg - away_missed_fg_pg,
    x3pa_rate_pg_diff  = home_x3pa_rate_pg - away_x3pa_rate_pg,
    expected_missed_fg = home_missed_fg_pg + away_missed_fg_pg
  )
```

# 4) Team Stats Per 100 Poss: pace-adjusted rebound strength

```{r}
p100_team_col   <- pick_col(team_100, c("team", "team_name", "teamName"))
p100_season_col <- pick_col(team_100, c("season"))
p100_orb_col    <- pick_col(team_100, c("orb_per_100_poss", "oreb_per_100_poss", "orb_per_100"))
p100_drb_col    <- pick_col(team_100, c("drb_per_100_poss", "dreb_per_100_poss", "drb_per_100"))
p100_trb_col    <- pick_col(team_100, c("trb_per_100_poss", "reb_per_100_poss", "trb_per_100"))
p100_fga_col    <- pick_col(team_100, c("fga_per_100_poss", "fga_per_100"))
p100_x3pa_col   <- pick_col(team_100, c("x3pa_per_100_poss", "3pa_per_100_poss", "x3pa_per_100"))

p100_team_col
p100_season_col
p100_orb_col
p100_drb_col
p100_trb_col
p100_fga_col
p100_x3pa_col
team_100_small <- team_100 %>%
  transmute(
    season = .data[[p100_season_col]],
    team   = .data[[p100_team_col]],
    orb_per100 = .data[[p100_orb_col]],
    drb_per100 = .data[[p100_drb_col]],
    trb_per100 = .data[[p100_trb_col]],
    fga_per100 = .data[[p100_fga_col]],
    x3pa_per100 = .data[[p100_x3pa_col]]
  ) %>%
  mutate(
    x3pa_rate_per100 = if_else(fga_per100 > 0, x3pa_per100 / fga_per100, NA_real_)
  ) %>%
  distinct()

home_100 <- team_100_small %>%
  rename_with(~ paste0("home_", .x), -c(season, team))

away_100 <- team_100_small %>%
  rename_with(~ paste0("away_", .x), -c(season, team))

games_master <- games_master %>%
  left_join(home_100, by = c("season", "home_team_full" = "team")) %>%
  left_join(away_100, by = c("season", "away_team_full" = "team")) %>%
  mutate(
    orb_per100_diff       = home_orb_per100 - away_orb_per100,
    drb_per100_diff       = home_drb_per100 - away_drb_per100,
    trb_per100_diff       = home_trb_per100 - away_trb_per100,
    x3pa_rate_per100_diff = home_x3pa_rate_per100 - away_x3pa_rate_per100,

    # matchup edges
    home_rebound_edge = home_orb_per100 - away_drb_per100,
    away_rebound_edge = away_orb_per100 - home_drb_per100,
    rebound_matchup_edge = home_rebound_edge - away_rebound_edge
  )
```

## Game-level rolling OREB opportunity features from TeamStatistics
```{r}
ts_game_id_col <- pick_col(team_stats, c("gameId"))
ts_team_col    <- pick_col(team_stats, c("teamName", "team"))
ts_home_col    <- pick_col(team_stats, c("home"))
ts_fg_col      <- pick_col(team_stats, c("fieldGoalsMade", "fg", "fgm"))
ts_fga_col     <- pick_col(team_stats, c("fieldGoalsAttempted", "fga"))
ts_x3p_col     <- pick_col(team_stats, c("threePointersMade", "x3p", "x3pm", "threePointFieldGoalsMade"))
ts_x3pa_col    <- pick_col(team_stats, c("threePointersAttempted", "x3pa", "threePointFieldGoalsAttempted"))
ts_oreb_col    <- pick_col(team_stats, c("reboundsOffensive", "oreb"))
ts_dreb_col    <- pick_col(team_stats, c("reboundsDefensive", "dreb"))

team_box_roll <- team_stats %>%
  transmute(
    game_id = as.character(.data[[ts_game_id_col]]),
    team    = .data[[ts_team_col]],
    home    = .data[[ts_home_col]],
    fg      = .data[[ts_fg_col]],
    fga     = .data[[ts_fga_col]],
    x3p     = .data[[ts_x3p_col]],
    x3pa    = .data[[ts_x3pa_col]],
    oreb    = .data[[ts_oreb_col]],
    dreb    = .data[[ts_dreb_col]]
  )

team_box_roll_home <- team_box_roll %>%
  filter(home == 1) %>%
  select(
    game_id,
    team,
    fg, fga, x3p, x3pa, oreb, dreb
  ) %>%
  left_join(games_master %>% select(game_id, date), by = "game_id") %>%
  mutate(
    pts_dummy = NA_real_,
    oreb_allowed = NA_real_
  )

team_box_roll_away <- team_box_roll %>%
  filter(home == 0) %>%
  select(
    game_id,
    team,
    fg, fga, x3p, x3pa, oreb, dreb
  ) %>%
  left_join(games_master %>% select(game_id, date), by = "game_id") %>%
  mutate(
    pts_dummy = NA_real_,
    oreb_allowed = NA_real_
  )

team_box_roll_long <- bind_rows(team_box_roll_home, team_box_roll_away) %>%
  mutate(
    date = as.Date(date),
    missed_fg = fga - fg,
    missed_3pt = x3pa - x3p,
    x3pa_rate_game = if_else(fga > 0, x3pa / fga, NA_real_)
  ) %>%
  arrange(team, date, game_id)

# build oreb_allowed from opponent same game
opp_oreb_lookup_home <- games_master %>%
  select(game_id, home_team, away_oreb)

opp_oreb_lookup_away <- games_master %>%
  select(game_id, away_team, home_oreb)

team_box_roll_long <- team_box_roll_long %>%
  left_join(opp_oreb_lookup_home, by = c("game_id", "team" = "home_team")) %>%
  left_join(opp_oreb_lookup_away, by = c("game_id", "team" = "away_team")) %>%
  mutate(
    oreb_allowed = coalesce(away_oreb, home_oreb)
  ) %>%
  select(-away_oreb, -home_oreb)

team_box_roll_long <- team_box_roll_long %>%
  group_by(team) %>%
  arrange(date, game_id, .by_group = TRUE) %>%
  mutate(
    last5_missed_fg = slide_dbl(lag(missed_fg), mean, .before = 4, .complete = TRUE),
    last10_missed_fg = slide_dbl(lag(missed_fg), mean, .before = 9, .complete = TRUE),

    last5_missed_3pt = slide_dbl(lag(missed_3pt), mean, .before = 4, .complete = TRUE),
    last10_missed_3pt = slide_dbl(lag(missed_3pt), mean, .before = 9, .complete = TRUE),

    last5_x3pa_rate = slide_dbl(lag(x3pa_rate_game), mean, .before = 4, .complete = TRUE),
    last10_x3pa_rate = slide_dbl(lag(x3pa_rate_game), mean, .before = 9, .complete = TRUE),

    last10_oreb = slide_dbl(lag(oreb), mean, .before = 9, .complete = TRUE),
    last5_oreb_allowed = slide_dbl(lag(oreb_allowed), mean, .before = 4, .complete = TRUE),
    last10_oreb_allowed = slide_dbl(lag(oreb_allowed), mean, .before = 9, .complete = TRUE)
  ) %>%
  ungroup()

home_oreb_roll <- team_box_roll_long %>%
  select(
    game_id,
    team,
    last5_missed_fg,
    last10_missed_fg,
    last5_missed_3pt,
    last10_missed_3pt,
    last5_x3pa_rate,
    last10_x3pa_rate,
    last10_oreb,
    last5_oreb_allowed,
    last10_oreb_allowed
  ) %>%
  rename(
    home_team = team,
    home_last5_missed_fg = last5_missed_fg,
    home_last10_missed_fg = last10_missed_fg,
    home_last5_missed_3pt = last5_missed_3pt,
    home_last10_missed_3pt = last10_missed_3pt,
    home_last5_x3pa_rate = last5_x3pa_rate,
    home_last10_x3pa_rate = last10_x3pa_rate,
    home_last10_oreb = last10_oreb,
    home_last5_oreb_allowed = last5_oreb_allowed,
    home_last10_oreb_allowed = last10_oreb_allowed
  )

away_oreb_roll <- team_box_roll_long %>%
  select(
    game_id,
    team,
    last5_missed_fg,
    last10_missed_fg,
    last5_missed_3pt,
    last10_missed_3pt,
    last5_x3pa_rate,
    last10_x3pa_rate,
    last10_oreb,
    last5_oreb_allowed,
    last10_oreb_allowed
  ) %>%
  rename(
    away_team = team,
    away_last5_missed_fg = last5_missed_fg,
    away_last10_missed_fg = last10_missed_fg,
    away_last5_missed_3pt = last5_missed_3pt,
    away_last10_missed_3pt = last10_missed_3pt,
    away_last5_x3pa_rate = last5_x3pa_rate,
    away_last10_x3pa_rate = last10_x3pa_rate,
    away_last10_oreb = last10_oreb,
    away_last5_oreb_allowed = last5_oreb_allowed,
    away_last10_oreb_allowed = last10_oreb_allowed
  )

games_master <- games_master %>%
  left_join(home_oreb_roll, by = c("game_id", "home_team")) %>%
  left_join(away_oreb_roll, by = c("game_id", "away_team")) %>%
  mutate(
    last5_missed_fg_diff = home_last5_missed_fg - away_last5_missed_fg,
    last10_missed_fg_diff = home_last10_missed_fg - away_last10_missed_fg,

    last5_missed_3pt_diff = home_last5_missed_3pt - away_last5_missed_3pt,
    last10_missed_3pt_diff = home_last10_missed_3pt - away_last10_missed_3pt,

    last5_x3pa_rate_diff = home_last5_x3pa_rate - away_last5_x3pa_rate,
    last10_x3pa_rate_diff = home_last10_x3pa_rate - away_last10_x3pa_rate,

    last10_oreb_diff = home_last10_oreb - away_last10_oreb,
    last5_oreb_allowed_diff = home_last5_oreb_allowed - away_last5_oreb_allowed,
    last10_oreb_allowed_diff = home_last10_oreb_allowed - away_last10_oreb_allowed,

    # matchup-style engineered features
    expected_oreb_pressure =
      (home_last5_missed_fg + away_last5_missed_fg) *
      ((home_orb_per100 + away_orb_per100) / 200),

    home_oreb_matchup = home_orb_per100 * away_last5_oreb_allowed,
    away_oreb_matchup = away_orb_per100 * home_last5_oreb_allowed,

    pace_rebound_pressure =
      ((home_last5_pace + away_last5_pace) / 2) *
      (home_last5_missed_fg + away_last5_missed_fg)
  )
```

# Rebuild modern data after adding new features
```{r}
games_master_modern <- games_master %>%
  filter(season >= 2010)
```

# 7) Rebuild final model dataset with new OREB predictors
```{r}
maya_team_history_home <- games_master %>%
  select(
    game_id,
    date,
    team = home_team,
    efg_pct = home_efg_pct,
    pace = home_pace,
    ft_rate = home_ft_rate,
    tov_pct = home_tov_pct
  )

maya_team_history_away <- games_master %>%
  select(
    game_id,
    date,
    team = away_team,
    efg_pct = away_efg_pct,
    pace = away_pace,
    ft_rate = away_ft_rate,
    tov_pct = away_tov_pct
  )

maya_team_history <- bind_rows(maya_team_history_home, maya_team_history_away) %>%
  mutate(date = as.Date(date)) %>%
  arrange(team, date, game_id) %>%
  group_by(team) %>%
  arrange(date, game_id, .by_group = TRUE) %>%
  mutate(
    shooting_env5   = slide_dbl(lag(efg_pct), mean, .before = 4, .complete = TRUE),
    pace_pressure5  = slide_dbl(lag(pace), mean, .before = 4, .complete = TRUE),
    free_throw_env5 = slide_dbl(lag(ft_rate), mean, .before = 4, .complete = TRUE),
    tov_env5        = slide_dbl(lag(tov_pct), mean, .before = 4, .complete = TRUE),
    blend_total_proj5 =
      (shooting_env5 + pace_pressure5 + free_throw_env5 + tov_env5) / 4
  ) %>%
  ungroup()

maya_home <- maya_team_history %>%
  select(
    game_id,
    team,
    shooting_env5,
    pace_pressure5,
    free_throw_env5,
    tov_env5,
    blend_total_proj5
  ) %>%
  rename(home_team = team)

maya_away <- maya_team_history %>%
  select(
    game_id,
    team,
    shooting_env5,
    pace_pressure5,
    free_throw_env5,
    tov_env5,
    blend_total_proj5
  ) %>%
  rename(
    away_team = team,
    away_shooting_env5 = shooting_env5,
    away_pace_pressure5 = pace_pressure5,
    away_free_throw_env5 = free_throw_env5,
    away_tov_env5 = tov_env5,
    away_blend_total_proj5 = blend_total_proj5
  )

games_master_modern <- games_master_modern %>%
  left_join(maya_home, by = c("game_id", "home_team")) %>%
  left_join(maya_away, by = c("game_id", "away_team")) %>%
  mutate(
    shooting_env5 = (shooting_env5 + away_shooting_env5) / 2,
    pace_pressure5 = (pace_pressure5 + away_pace_pressure5) / 2,
    free_throw_env5 = (free_throw_env5 + away_free_throw_env5) / 2,
    tov_env5 = (tov_env5 + away_tov_env5) / 2,
    blend_total_proj5 = (blend_total_proj5 + away_blend_total_proj5) / 2
  )


```


# 8) Final cleanup

```{r}
games_master_modern <- games_master_modern %>%
  filter(!is.na(home_last5_pts), !is.na(away_last5_pts)) %>%
  mutate(
    pace_efficiency = pace_pressure5 * shooting_env5,
    extra_possession_pressure = expected_oreb_pressure + rebound_matchup_edge,
    expected_possessions = pace_pressure5 + pace_rebound_pressure,
    scoring_efficiency = (shooting_env5 + free_throw_env5) / 2
  )

model_data <- games_master_modern %>%
  mutate(
    home_team = coalesce(home_team_full, home_team),
    away_team = coalesce(away_team_full, away_team)
  ) %>%
  select(
    game_id,
    date,
    season,
    home_team,
    away_team,
    spread,
    total,
    oreb,

    opp_pts_allowed_diff,
    opp_orb_allowed_diff,
    opp_trb_allowed_diff,
    opp_ast_allowed_diff,
    opp_tov_forced_diff,

    off_rtg_diff,
    def_rtg_diff,
    net_rtg_diff,
    efg_pct_diff,
    tov_pct_diff,
    orb_pct_diff,
    ft_rate_diff,
    last5_off_rtg_diff,
    last5_def_rtg_diff,
    last5_net_rtg_diff,
    last5_pace_diff,

    last5_pts_diff,
    last5_pts_allowed_diff,
    last5_oreb_diff,

    rest_diff,
    home_back_to_back,
    away_back_to_back,

    shooting_env5,
    pace_pressure5,
    free_throw_env5,
    tov_env5,
    blend_total_proj5,

    orb_pg_diff,
    drb_pg_diff,
    trb_pg_diff,
    missed_fg_pg_diff,
    x3pa_rate_pg_diff,
    expected_missed_fg,

    orb_per100_diff,
    drb_per100_diff,
    trb_per100_diff,
    x3pa_rate_per100_diff,
    rebound_matchup_edge,

    last5_missed_fg_diff,
    last10_missed_fg_diff,
    last5_missed_3pt_diff,
    last10_missed_3pt_diff,
    last5_x3pa_rate_diff,
    last10_x3pa_rate_diff,
    last10_oreb_diff,
    last5_oreb_allowed_diff,
    last10_oreb_allowed_diff,
    expected_oreb_pressure,
    home_oreb_matchup,
    away_oreb_matchup,
    pace_rebound_pressure,

    pace_efficiency,
    extra_possession_pressure,
    expected_possessions,
    scoring_efficiency
  ) %>%
  replace_na(list(
    rest_diff = 0,
    home_back_to_back = 0,
    away_back_to_back = 0
  )) %>%
  drop_na(game_id, date, spread, total, oreb)

model_data_v2 <- model_data %>%
  mutate(date = as.Date(date)) %>%
  filter(month(date) %in% c(10, 11, 12, 1, 2, 3, 4)) %>%
  rename(game_date = date)
```

Save
```{r}
write_csv(model_data_v2, "Data/FINAL_MODEL_READY.csv")
```



